---
title: "Personal Activity Monitoring Device Patterns Assesment"
author: "Abiyu Giday"
date: "September 18, 2015"
output: 
         html_document:
             toc: true
---

#### _Table of Content_
1. Overview
2. Q-1 Loading and preprocessing the data
   + Process/transform the data (if necessary) into a format suitable for your analysis
3. Q-2 What is mean total number of steps taken per day?
   + Calculate the total number of steps taken per day
   + Make a histogram of the total number of steps taken each day
   + Calculate and report the mean and median of the total number of steps taken per day
4. Q-3 What is the average daily activity pattern?
   + Make a time series plot on the average number of steps taken
   + Which 5-minute interval contains the maximum number of steps?
5. Q-4 Imputing missing values
   + Calculate and report the total number of missing values in the dataset
   + Impute missing values in the dataset
   + Create a new dataset that is equal to the original dataset but with the missing data filled in.
   + Make a histogram of the total number of steps taken each day and Calculate and report the + mean and median total number of steps taken per day.
6. Q-5 Are there differences in activity patterns between weekdays and weekends?
   + Preparing the data table for weekday/weekend activty pattern comparison
   + Create a new factor variable with two levels – “weekday” and “weekend”
   + Make a time series plot the average number of steps taken, averaged across all weekday days or weekend days
7. What are the weekly Distribution of steps taken for the two months?
   + Daily steps bar chart
8. Over all observation

# Overview
<img src="http://www.ericselectronics.com/wp-content/uploads/2015/06/HAMSWAN-Smart-Wrist-Watch-Bluetooth-40-Fitness-Tracker-Health-Smartwatch-with-Heart-Rate-Monitor-Phone-Watches-NFC-Function-for-IOS-Android-Iphone-Samsung-HTC-Smartphones-White-0.jpg" align="right" width="35%" height="35%" />
This analysis makes use of data from a personal activity monitoring device, such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take   measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there 
is a lack of statistical methods and software for processing and interpreting the data. This script Calculate the total number of steps taken per day, calculate and report the mean and median of the total number of steps, plots histogram of the total number of steps, shows average daily activity pattern,  method that impute missing values from the data, and Make a panel plot containing a time series plots. 

## Q-1 Loading and preprocessing the data

```{r}
# Download and save the file in data directory
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "./data/Factivity.zip", method = "curl")
unzip("./data/Factivity.zip")
```

### Process/transform the data (if necessary) into a format suitable for your analysis
```{r, echo=TRUE}
library(dplyr)

actvty <- read.csv("activity.csv") #data frame in active directory
actvty <- tbl_df(actvty) # convert df to dt
actvty1 <- na.omit(actvty)  # remove missing rows from the data frame
actvty1 <- select(actvty1, date, interval, steps)       # rearrange the data columns
actvty1 <- actvty1 %>%
        group_by(date) %>% 
                summarise_each(funs(sum), steps) # use dplyr to sum steps each day
str(actvty1)
head(actvty1)

```

##Q-2 What is mean total number of steps taken per day?

```{r, echo=TRUE}
TotalSteps <- actvty1$steps 
summary(TotalSteps) 
```


### Calculate the total number of steps taken per day
```{r, echo=TRUE}
sum(actvty1$steps)
```
The total number steps take is `r sum(actvty1$steps)`.

### Make a histogram of the total number of steps taken each day
#### Figure 1: A histogram for 10/12 & 11/2
```{r, Fig-1dayTotalSteps, echo=TRUE}
actvtyHist <- hist(actvty1$steps,  breaks = 15, freq = T, col = "yellow", 
                   main = "Histogram for total number of Steps taken per day",
                   xlab = "Daily step count", ylab ="Frequency")
rug(actvty1$steps) # rug under the histogram to show step concentration
abline(v = mean(actvty1$steps), col = "red", lwd = 4)
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r, echo=TRUE}
TotalSteps <- actvty1$steps 
summary(TotalSteps) 
```

##Q-3 What is the average daily activity pattern?
### Make a time series plot on the average number of steps taken

Loading and preprocessing the data for the time series plot
```{r, echo=TRUE}
library(ggplot2)

str(actvty)
actvty$date <- as.Date(actvty$date)
actvty2sum <- actvty %>% 
        group_by(date) %>% 
                    summarise_each(funs(sum), steps) # sum the steps for each day
actvty2sum <- na.omit(actvty2sum)      #remove rows with unavilable data - NA
head(actvty2sum)
```

####Figure 2: A Time series plot for the total Steps Across all days
```{r, Fig-2TSstepsAcrossDays, echo=TRUE}
qplot(steps, date, data = actvty2sum, geom = c("line", "smooth"), method ="lm") + 
        ggtitle("Average total steps Across all days") + 
        labs( x = "Steps", y = "Days") +
        theme(plot.title = element_text(lineheight=.8, face="bold")) 
```

### Which 5-minute interval contains the maximum number of steps?
 
```{r, echo=TRUE}
 summary(actvty)
 filter(actvty, steps ==806)
```

##Q-4 Imputing missing values
To impute value for missing data a numberof model/techhniques. For this excersie we are using the **Last value carried forward _(LOCF)_** technique.  The last observed value is used to fill in missing values in subsequent observations, this method assums that the most recent observation is the best guess for subsequent missing values. We are using the R **_ZOO_** package's **na.locf** function.
### Prepare the data 
```{r, echo=TRUE}
actvty3 <- read.csv("activity.csv")                #data frame in active directory

actvty3 <- actvty3 %>%
             group_by(date) %>% 
                summarise_each(funs(sum), steps)   # sum of  steps each day

```
### Calculate and report the total number of missing values in the dataset
```{r, echo = TRUE}
sum(is.na(actvty3)) # how many days of misseang value
mean(is.na(actvty3)) # 4.4% NA
```
#### Does the presence of missing days introduce bias?
In order to answer the imapct of bias after imputing data, the following steps in the _zoo_ package na.locf techinque is utlized as follows.

### Impute missing values in the dataset
```{r, echo=TRUE}
library(zoo)

actvty3 <- actvty3 %>%
             group_by(date) %>% 
                summarise_each(funs(sum), steps)   # sum of  steps each day

# change the var type back to Date/numeric
actvty3$steps <- as.numeric(actvty3$steps)
actvty3$date <- as.Date(actvty3$date)
summary(actvty3$steps)

actvty33 <- na.locf(actvty3, na.rm = FALSE) #replace NA with the Last Value from the top down
actvty333 <- na.locf(actvty33, na.rm = FALSE, fromLast = TRUE) # NA replace from the last data point to complete the cricle

# change the var type back to Date/numeric
actvty333$steps <- as.numeric(actvty333$steps)
actvty333$date <- as.Date(actvty333$date)
```
#### Confirming Imputed value remove all NA
```{r echo=TRUE}
sum(is.na(actvty333)) # how many days of misseang value
mean(is.na(actvty333))
ImpTotlSteps <- actvty3$steps
summary(ImpTotlSteps)
```
### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo=TRUE}
actvty333 <- tbl_df(actvty333)
str(actvty333)
sum(is.na(actvty333))
head(actvty333)
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.
#### Prepare the data table for comarative histogram plot
``{r, echo=TRUE}
#change variable data types 
actvty333$date <- as.Date(actvty333$date)
actvty333$interval <- as.numeric(actvty333$interval)
actvty333$steps <- as.numeric(actvty333$steps)
actvty333sum <- actvty333 %>%
                  group_by(date) %>% 
                     summarise_each(funs(sum), steps) # sum of  steps each day



#### What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r, echo= TRUE}
AddSteps = sum(actvty333$steps) - sum(actvty3$steps, na.rm = TRUE)
T =  AddSteps / sum(actvty3$steps, na.rm = TRUE) * 100
```
The imputed data with locf technique added `r AddSteps` additional steps.Which is `r T` % more, and this could add a slight bias in the analysis. 

##### Side by side summary comparison shows in this table


     Summary  |  Msng NA  | Imputed NA|
     -------- | ----------|-----------| 
     Min      |   41      |    41     |   
     Median   |   10760   |    10570  |
     Mean     |   10770   |    10300  |
     Max      |   21990   |    21990  |
     Total    |   570608  |    628555 |

     
The imputed data with locf technique added `r AddSteps` many more steps. Which is `r T` % more. That is statstically significant could introduce bias in the analysis. 
     
#### Figure 3: Side by side comparison of total steps with missing and imputed data.
```{r, Fig-3compareNAvsimputed, echo=TRUE, fig.width=8, fig.height=5}
par(mfrow = c(1,2))

actvtyHist <- hist(actvty1$steps,  breaks = 15, freq = T, col = "yellow", 
                   main = "Total steps with missing data removd",
                   xlab = "Daily step count", ylab ="Frequency")
                   rug(actvty1$steps)
                   abline(v = mean(actvty1$steps), col = "red", lwd = 4)

actvtyHist3 <- hist(actvty333$steps,  breaks = 15, freq = T, col = "Green", 
                    main = "Total steps with missing data imputed",
                    xlab = "Daily step count", ylab ="Frequency")
                    rug(actvty333$steps)
                    abline(v = mean(actvty333$steps), col = "red", lwd = 4)
```



##Q-5 Are there differences in activity patterns between weekdays and weekends?
### Preparing the data table for weekday/weekend activty pattern comparison
```{r, echo=TRUE}
actvtyWK <- actvty333  %>% mutate( DaysOfWk = weekdays(as.Date(date))) # Add coresponding weekday
str(actvtyWK)
head(actvtyWK, 10)
```

### Create a new factor variable with two levels – “weekday” and “weekend” 
```{r, echo=TRUE}
actvtyWKD <- actvtyWK %>%  mutate(TypeOfDay = ifelse(DaysOfWk == "Saturday" | DaysOfWk =="Sunday", "weekend", "weekday" )) # add weekday & weekend variable 

actvtyWKD[actvtyWKD == 0] <- NA  # convert 0 value data to NA
actvtyWKD1 <- na.locf(actvtyWKD, na.rm = FALSE) #replace NA with the Last Value from the top down
actvtyWKD2 <- na.locf(actvtyWKD1, na.rm = FALSE, fromLast = TRUE) # NA replace from the last data point to complete the cricle

# #change variable data types 
actvtyWKD2$TypeOfDay <- as.factor(actvtyWKD2$TypeOfDay)
actvtyWKD2$DaysOfWk <- as.factor(actvtyWKD2$DaysOfWk) 
actvtyWKD2$steps <- as.numeric(actvtyWKD2$steps)
actvtyWKD2$date <- as.Date(actvtyWKD2$date)

str(actvtyWKD2)
head(actvtyWKD2, 10)



```

###


### Make a time series plot the average number of steps taken, averaged across all weekday days or weekend days
#### Figure 4: Weekdays Vs Weekend time series plot 
```{r, Fig-4weekdayVSweekend, echo=TRUE}
qplot(steps, date, data = actvtyWKD2, geom = c("line", "smooth"), method ="lm",facets = .~TypeOfDay) + 
        ggtitle("Average total steps on Weekdays and Weekends") + 
        labs( x = "Steps", y = "days") +
        theme(plot.title = element_text(lineheight=.8, face="bold"))
```

## Q-6 What are the weekly Distribution of steps taken for the two months?
```{r, echo= TRUE}
actvtyWKD2

actvtyPerDay <- actvtyWKD2 %>%
                  group_by(DaysOfWk) %>% 
                     summarise_each(funs(sum), steps) # sum of  steps each day
Dailyactv <- arrange(actvtyPerDay, desc(steps))
Dailyactv

```
### Figure 5: Daily steps bar chart
```{r, Fig-5dailyActvty, echo = TRUE}
ggplot(Dailyactv, aes(x = factor(DaysOfWk), y = steps, color = DaysOfWk)) + geom_bar(stat = "identity") + ggtitle("Steps per day distribution") + xlab("Days of the week") 
```

# Over all observation
-  For the observed activty data, the steps tend to increase from Monday to Friday. On Weekend About the same number of steps taken on saturday and sunday.
- The imputed data with locf technique added `r AddSteps` additional steps. Which is `r T` % more. This could add a slight bias in the analysis, however the model factors in somemarigin of error.  
-  November 27, the 615th interval contains the maximum number of steps.
- The total number steps take not including the missing data is `r sum(actvty1$steps)`.
The subject the activty monitor data analyzed walked less at the beggining of the week, increasing  activity towards the end of the week.  
- The most steps are taken on Wednesdays, with 26% more steps compare to least active day Tuesdays.



